---
title: "Untitled"
output: html_document
date: "2024-04-17"
---

```{r}
library(ggplot2)
library(reshape2)
library(MASS)
library(nortest)

set.seed(42) 

```

# read data
```{r}
# red_wine <- read.csv("winequality-red.csv", sep = ";") 
# white_wine <- read.csv("winequality-white.csv", sep = ";")
# 
# red_wine$wine_type <- 'red'
# white_wine$wine_type <- 'white'
# 
# red_wine_sample <- red_wine[sample(nrow(red_wine), 25), ]
# white_wine_sample <- white_wine[sample(nrow(white_wine), 25), ]
# 
# wine <- rbind(red_wine_sample, white_wine_sample)
# write.csv(wine, "wine.csv", row.names = FALSE)

wine <- read.csv("wine.csv") 
```



# Get to know data
```{r}
# dimension of data
dim(winedata) # 6497 rows x 13 variables

# features in data
names(winedata) 

```

# Categorical data visualization
```{r}
# Comments
# Most common wine type = white wine
# Most common quality grade = 6, 5, & 7

# Get the frequency table
quality_table <- table(winedata$quality)
type_table <- table(winedata$wine_type)

# View the frequency table
print(quality_table)
print(type_table)


# barplot
winedata$quality <- factor(winedata$quality)
ggplot(winedata, aes(x = quality)) + geom_bar(fill = "skyblue") + 
  labs(x = "quality", y = "count")

winedata$wine_type <- factor(winedata$wine_type)
ggplot(winedata, aes(x = wine_type)) + geom_bar(fill = "skyblue") + 
  labs(x = "type of wine", y = "count")
```

# Quantitative data visualization 

```{r}
# Get summary of all variables
summary(winedata[,1:11])
```

## histogram
```{r}
ggplot(winedata, aes(x = fixed.acidity)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "fixed acidity", y = "count") + theme_minimal()

ggplot(winedata, aes(x = volatile.acidity)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "volatile acidity", y = "count") + theme_minimal()

ggplot(winedata, aes(x = citric.acid)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "citric acid", y = "count") + theme_minimal()

ggplot(winedata, aes(x = residual.sugar)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "residual sugar", y = "count") + theme_minimal()

ggplot(winedata, aes(x = chlorides)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "chlorides", y = "count") + theme_minimal()

ggplot(winedata, aes(x = free.sulfur.dioxide)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "free sulfur dioxide", y = "count") + theme_minimal()

ggplot(winedata, aes(x = total.sulfur.dioxide)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "total sulfur dioxide", y = "count") + theme_minimal()

ggplot(winedata, aes(x = density)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = " ", y = "count") + theme_minimal()

ggplot(winedata, aes(x = density)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "density", y = "count") + theme_minimal()

ggplot(winedata, aes(x = pH)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "pH", y = "count") + theme_minimal()

ggplot(winedata, aes(x = sulphates)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "sulphates", y = "count") + theme_minimal()

ggplot(winedata, aes(x = alcohol)) +
  geom_histogram(color = "white", fill = "skyblue") +
  labs(x = "alcohol", y = "count") + theme_minimal()
```

## density plots
```{r}
ggplot(winedata, aes(x = fixed.acidity)) +
  geom_density() +
  labs(title = "Density Plot", x = "fixed acidity") +
  theme_minimal()
```

## boxplot
```{r}
ggplot(winedata, aes(x = fixed.acidity)) + geom_boxplot() +
  labs(title = "Box Plot of fixed acidity", x = "fixed acidity") +
  theme_minimal()

ggplot(winedata, aes(x = volatile.acidity)) + geom_boxplot() +
  labs(title = "Box Plot of volatile.acidity", x = "volatile.acidity") + theme_minimal()

ggplot(winedata, aes(x = citric.acid)) + geom_boxplot() +
  labs(title = "Box Plot of citric.acid", x = "citric.acid") + theme_minimal()

ggplot(winedata, aes(x = residual.sugar)) + geom_boxplot() +
  labs(title = "Box Plot of residual.sugar", x = "residual.sugar") + theme_minimal()

ggplot(winedata, aes(x = chlorides)) + geom_boxplot() +
  labs(title = "Box Plot of chlorides", x = "chlorides") + theme_minimal()

ggplot(winedata, aes(x = free.sulfur.dioxide)) + geom_boxplot() +
  labs(title = "Box Plot of free.sulfur.dioxide", x = "free.sulfur.dioxide") + theme_minimal()

ggplot(winedata, aes(x = total.sulfur.dioxide)) + geom_boxplot() +
  labs(title = "Box Plot of total sulfur dioxide", x = "total.sulfur.dioxide") + theme_minimal()

ggplot(winedata, aes(x = density)) + geom_boxplot() +
  labs(title = "Box Plot of density", x = "density") + theme_minimal()

ggplot(winedata, aes(x = pH)) + geom_boxplot() +
  labs(title = "Box Plot of pH", x = "pH") + theme_minimal()

ggplot(winedata, aes(x = sulphates)) + geom_boxplot() +
  labs(title = "Box Plot of sulphates", x = "sulphates") + theme_minimal()

ggplot(winedata, aes(x = alcohol)) + geom_boxplot() +
  labs(title = "Box Plot of alcohol", x = "alcohol") + theme_minimal()# Correlation matrix
```

# correlation matrix
```{r}
subset_winedata <- winedata[, 1:11]
cor_matrix <- cor(subset_winedata)

# Melt the correlation matrix for ggplot2
cor_melted <- melt(cor_matrix)

# Create the heatmap
heatmap <- ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() + # This creates the heatmap squares
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Correlation") +
  geom_text(aes(label = sprintf("%.2f", value)), color = "black", size = 3) +
  theme_minimal() + # Clean theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x axis labels
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

# Print the heatmap
print(heatmap)

```

# normality check
```{r}
check_normality <- function(data, var) {
  test_result <- lillie.test(data[[var]])
  is_normal <- if (test_result$p.value < 0.05) "not normal" else "normal ***"
  cat(var, "has p-value:", test_result$p.value,  "is", is_normal,"\n")
  return(test_result$p.value)
}

variables <- names(wine)[1:11]
for (var in variables) {
  p_value <- check_normality(wine, var)
}
```


```{r}
# variables <- names(wine[,c(1,2,4,5)])
# for (var in variables) {
#   hist(wine[[var]], main=paste("Histogram of", var), xlab=var, col="lightblue", border="black")
#   # Adds a normal curve for reference
#   mean_val <- mean(wine[[var]], na.rm = TRUE)
#   sd_val <- sd(wine[[var]], na.rm = TRUE)
#   curve(dnorm(x, mean=mean_val, sd=sd_val), add=TRUE, col="red")
# }

```

## boxcox transformation
```{r}
apply_boxcox_save_check <- function(data, var) {
  if (any(data[[var]] <= 0)) {
    offset <- abs(min(data[[var]], na.rm = TRUE)) + 1
    data[[var]] <- data[[var]] + offset
  }
  
  bc_transform <- boxcox(data[[var]] ~ 1, lambda = seq(-5, 5, by = 0.1))
  lambda_optimal <- bc_transform$x[which.max(bc_transform$y)]
  
  if (lambda_optimal != 0) {
    transformed_var <- (data[[var]]^lambda_optimal - 1) / lambda_optimal
  } else {
    transformed_var <- log(data[[var]])  # Use log transformation when lambda is zero
  }
  
  test_result <- shapiro.test(transformed_var)
  normality_status <- if (test_result$p.value < 0.05) "not normal" else "normal"
  
  if (normality_status == "normal") {
    data[[var]] <- transformed_var
  }
  
  cat("P-value:", test_result$p.value, var, "with λ =", lambda_optimal, normality_status, "\n")
  
  return(data)
}

variables <- names(wine[, c(1, 2, 4, 5)])
wine1 <- wine  # Initialize wine1 as a copy of wine

for (var in variables) {
  wine1 <- apply_boxcox_save_check(wine1, var)
  cat("\n")
}
```
```{r}
check_normality <- function(data, var) {
  test_result <- lillie.test(data[[var]])
  is_normal <- if (test_result$p.value < 0.05) "not normal" else "normal ***"
  cat(var, "has p-value:", test_result$p.value,  "is", is_normal,"\n")
  return(test_result$p.value)
}

variables <- names(wine1)[1:11]
for (var in variables) {
  p_value <- check_normality(wine1, var)
}
write.csv(wine1, "wine1.csv", row.names = FALSE)
```


## other tranformation
```{r}
# # Function to apply various transformations and check normality
# transform_check_normality <- function(data, var) {
#   
#   if (any(data[[var]] <= 0)) {
#     # Add 1 to all elements to make them positive
#     data[[var]] <- data[[var]] + 1
#   }
#   
#   transformations <- list(
#     "log" = log(data[[var]]),
#     "sqrt" = sqrt(data[[var]]),
#     "inverse" = 1 / data[[var]],
#     "exp" = exp(data[[var]]),
#     "rank" = rank(data[[var]]) 
#   )
#   
#   results <- sapply(transformations, function(x) {
#     if (any(is.na(x))) return(c(p_value = NA, is_normal = NA))
#     test_result <- shapiro.test(x)
#     return(c(p_value = test_result$p.value, is_normal = ifelse(test_result$p.value < 0.05, "not normal", "normal")))
#   })
#   
#   return(results)
# }
# 
# 
# variables <- names(wine[,c(1,2,4,5)])
# results_list <- list()
# for (var in variables) {
#     result <- transform_check_normality(wine, var) 
#     results_list[[var]] <- result  
# }
# 
# print(results_list)

```
# Check for multivariate normality

# Models

## a. Data Reduction or Structural Simplification (Jarrad)
### PCA
### FA

## b. Grouping or Discrimination
### LDA (chapter 11) (Maddie)
### QDA (chapter 11) (Maddie)
### Cluster Analysis (Jiali)


## c. Dependence Investigation (Jiali)
### Correlation Analysis
### Canonical Correlation Analysis (chapter 10)

## d. Prediction or classification (Maddie)
## Logistic regression (chapter 11)
### Multivariate Regression

## e. Hypothesis construction and testing (Jarrad)
### one-way MANOVA




```{r}

```

