---
title: "Untitled"
output: html_document
date: "2024-04-17"
---

```{r}
library(ggplot2)
library(reshape2)
library(MASS)
library(nortest)

set.seed(42) 
```
# Introduction
## Back ground 
Wine is a product that can vary in price and in quality ; some varieties are akin to cheap consummers goods while others are luxury products. In that context, it can be hard for consummers to identify which are more appropriate for certain occasions, or which are worth spending more or less money on ; similarly, for producers, the task of setting up a price can be made harder. On both sides, certifications are very important and thus, there is a real need to build trustworthy models to evaluate wines quality: it would bring clarity to consumers and recognition to producers. Moreover, by identifying the most important factors, the latter could turn their focus on these aspects and find more efficient ways to improve their wine's rating.

## Variables

Our project is designed to harness the rich insights of a comprehensive public dataset, featuring detailed attributes and quality assessments of wine samples from Minho, a renowned wine-producing region in north-west Portugal. The dataset encompasses an extensive period of data collection, from May 2004 to February 2007, and includes an impressive total of 6,498 records related to both red and white wine variants, dissected across 11 key attributes. This meticulous compilation offers an unparalleled opportunity to delve into the nuances that distinguish wine quality, presenting a foundation for robust analysis and potential advancements in wine science. The dataset provides a granular look at the chemical composition and sensory attributes of wine, each variable casting light on its potential impact on overall quality. These attributes include:

    Fixed Acidity ($X_1$, numeric): Reflects the concentration of nonvolatile acids (such as tartaric acid) that remain fixed during winemaking and contribute to the wine's structure.
    Volatile Acidity ($X_2$, numeric): Measures the volatile acids (like acetic acid), where excessive levels can mar wine with an undesirable vinegar taste.
    Citric Acid ($X_3$, numeric): Although present in small amounts, citric acid can enhance the wine's freshness and flavor profile.
    Residual Sugar ($X_4$, numeric): Indicates the sugar level post-fermentation, affecting sweetness.
    Chlorides ($X_5$, numeric): The measure of salt content in wine, impacting taste.
    Free Sulfur Dioxide ($X_6$, numeric): Represents the portion of sulfur dioxide not bound to other molecules, crucial for preserving wine's freshness and inhibiting microbial growth.
    Total Sulfur Dioxide ($X_7$, numeric): The total concentration of sulfur dioxide, encompassing both free and bound forms, essential for wine longevity.
    Density ($X_8$, numeric): Reflects the wine's density, which correlates with its alcohol and sugar content.
    pH ($X_9$, numeric): A vital indicator of acidity, influencing taste, color, and stability.
    Sulphates ($X_{10}$, numeric): Pertains to added sulphates like potassium sulphate, affecting microbial stability and antioxidant properties.
    Alcohol ($X_{11}$, numeric): The alcohol percentage, directly influencing flavor and body.
    Quality ($X_{12}$, discrete): An assessment of wine quality on a scale from 0 to 10, based on sensory evaluation.
    Wine Type ($X_{13}$, categorical): Distinguishes between red and white wine variants.

By exploring these attributes, our project aims not only to dissect the complex interplay of factors influencing wine quality but also to contribute valuable insights to the wine industry, enhancing our understanding of what makes a wine stand out.

```{r}
red_wine <- read.csv("winequality-red.csv", sep = ";") 
white_wine <- read.csv("winequality-white.csv", sep = ";")

red_wine$wine_type <- 'red'
white_wine$wine_type <- 'white'

wine <- rbind(red_wine, white_wine)
write.csv(wine, "wine.csv", row.names = FALSE)
```

## variables
Our project is designed to harness the rich insights of a comprehensive public dataset, featuring detailed attributes and quality assessments of wine samples from Minho, a renowned wine-producing region in north-west Portugal. The dataset encompasses an extensive period of data collection, from May 2004 to February 2007, and includes an impressive total of 6,498 records related to both red and white wine variants, dissected across 11 key attributes. This meticulous compilation offers an unparalleled opportunity to delve into the nuances that distinguish wine quality, presenting a foundation for robust analysis and potential advancements in wine science.
The dataset provides a granular look at the chemical composition and sensory attributes of wine, each variable casting light on its potential impact on overall quality. These attributes include:
- **Fixed Acidity** (numeric): Reflects the concentration of nonvolatile acids (such as tartaric acid) that remain fixed during winemaking and contribute to the wine's structure.
- **Volatile Acidity** (numeric): Measures the volatile acids (like acetic acid), where excessive levels can mar wine with an undesirable vinegar taste.
- **Citric Acid** (numeric): Although present in small amounts, citric acid can enhance the wine's freshness and flavor profile.
- **Residual Sugar** (numeric): Indicates the sugar level post-fermentation, affecting sweetness.
- **Chlorides** (numeric): The measure of salt content in wine, impacting taste.
- **Free Sulfur Dioxide** (numeric): Represents the portion of sulfur dioxide not bound to other molecules, crucial for preserving wine's freshness and inhibiting microbial growth.
- **Total Sulfur Dioxide** (numeric): The total concentration of sulfur dioxide, encompassing both free and bound forms, essential for wine longevity.
- **Density** (numeric): Reflects the wine's density, which correlates with its alcohol and sugar content.
- **pH** (numeric): A vital indicator of acidity, influencing taste, color, and stability.
- **Sulphates** (numeric): Pertains to added sulphates like potassium sulphate, affecting microbial stability and antioxidant properties.
- **Alcohol** (numeric): The alcohol percentage, directly influencing flavor and body.
- **Wine Type** (categorical): Distinguishes between red and white wine variants.
- **Quality** (discrete): An assessment of wine quality on a scale from 0 to 10, based on sensory evaluation.

By exploring these attributes, our project aims not only to dissect the complex interplay of factors influencing wine quality but also to contribute valuable insights to the wine industry, enhancing our understanding of what makes a wine stand out.

# EDA of Data
# Data Normalization
## 1. univariate normality
```{r}
check_normality <- function(data, var) {
  test_result <- lillie.test(data[[var]])
  is_normal <- if (test_result$p.value < 0.05) "not normal" else "normal ***"
  cat(var, "has p-value:", test_result$p.value,  "is", is_normal,"\n")
  return(test_result$p.value)
}

variables <- names(wine)[1:11]
for (var in variables) {
  p_value <- check_normality(wine, var)
}
```


```{r}
# variables <- names(wine[,c(1,2,4,5)])
# for (var in variables) {
#   hist(wine[[var]], main=paste("Histogram of", var), xlab=var, col="lightblue", border="black")
#   # Adds a normal curve for reference
#   mean_val <- mean(wine[[var]], na.rm = TRUE)
#   sd_val <- sd(wine[[var]], na.rm = TRUE)
#   curve(dnorm(x, mean=mean_val, sd=sd_val), add=TRUE, col="red")
# }

```

## boxcox transformation
```{r}
apply_boxcox_save_check <- function(data, var) {
  if (any(data[[var]] <= 0)) {
    offset <- abs(min(data[[var]], na.rm = TRUE)) + 1
    data[[var]] <- data[[var]] + offset
  }
  
  bc_transform <- boxcox(data[[var]] ~ 1, lambda = seq(-5, 5, by = 0.1))
  lambda_optimal <- bc_transform$x[which.max(bc_transform$y)]
  
  if (lambda_optimal != 0) {
    transformed_var <- (data[[var]]^lambda_optimal - 1) / lambda_optimal
  } else {
    transformed_var <- log(data[[var]])  # Use log transformation when lambda is zero
  }
  
  test_result <- shapiro.test(transformed_var)
  normality_status <- if (test_result$p.value < 0.05) "not normal" else "normal"
  
  if (normality_status == "normal") {
    data[[var]] <- transformed_var
  }
  
  cat("P-value:", test_result$p.value, var, "with λ =", lambda_optimal, normality_status, "\n")
  
  return(data)
}

variables <- names(wine[, c(1, 2, 4, 5)])
wine1 <- wine  # Initialize wine1 as a copy of wine

wine1 <- read.csv('wine1.csv')
for (var in variables) {
  wine1 <- apply_boxcox_save_check(wine1, var)
  cat("\n")
}
```
```{r}
check_normality <- function(data, var) {
  test_result <- lillie.test(data[[var]])
  is_normal <- if (test_result$p.value < 0.05) "not normal" else "normal ***"
  cat(var, "has p-value:", test_result$p.value,  "is", is_normal,"\n")
  return(test_result$p.value)
}

variables <- names(wine1)[1:11]
for (var in variables) {
  p_value <- check_normality(wine1, var)
}
write.csv(wine1, "wine1.csv", row.names = FALSE)
```


## other tranformation
```{r}
# # Function to apply various transformations and check normality
# transform_check_normality <- function(data, var) {
#   
#   if (any(data[[var]] <= 0)) {
#     # Add 1 to all elements to make them positive
#     data[[var]] <- data[[var]] + 1
#   }
#   
#   transformations <- list(
#     "log" = log(data[[var]]),
#     "sqrt" = sqrt(data[[var]]),
#     "inverse" = 1 / data[[var]],
#     "exp" = exp(data[[var]]),
#     "rank" = rank(data[[var]]) 
#   )
#   
#   results <- sapply(transformations, function(x) {
#     if (any(is.na(x))) return(c(p_value = NA, is_normal = NA))
#     test_result <- shapiro.test(x)
#     return(c(p_value = test_result$p.value, is_normal = ifelse(test_result$p.value < 0.05, "not normal", "normal")))
#   })
#   
#   return(results)
# }
# 
# 
# variables <- names(wine[,c(1,2,4,5)])
# results_list <- list()
# for (var in variables) {
#     result <- transform_check_normality(wine, var) 
#     results_list[[var]] <- result  
# }
# 
# print(results_list)

```

## Check for multivariate normality

# Models

## a. Data Reduction or Structural Simplification (Jarrad)
### PCA
### FA

## b. Grouping or Discrimination

### Covariance matrix comparison
```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}
g = 2
p = 11

Sp = matrix(0, nrow = p, ncol = p)
S = list()
X_mean = list()
total_df = 0
total_inv_df = 0
wt_sum_logdet_cov = 0

for (i in 1:g) {
  if (g==1){group_data = red_wine[,c(-12, -13)]}
  if (g==2){group_data = white_wine[,c(-12, -13)]}
  
  m = dim(group_data)[1]
  X_mean[[i]] = colMeans(group_data)
  S[[i]] = cov(group_data)
  
  total_df = total_df + (m - 1)
  total_inv_df = total_inv_df + 1 / (m - 1)
  
  Sp = Sp + (m - 1) * S[[i]]
  wt_sum_logdet_cov = wt_sum_logdet_cov + (m - 1) * log(det(S[[i]]))
}

Sp = Sp / total_df
M = total_df * log(det(Sp)) - wt_sum_logdet_cov
u = (total_inv_df - 1 / total_df) * ((2 * p^2 + 3 * p - 1) / (6 * (p + 1) * (g - 1)))

C_obs = (1 - u) * M
nu = p * (p + 1) * (g - 1) / 2
p_value = 1 - pchisq(C_obs, df = nu)

if (p_value < 0.05){
  print(c("p-value =", p_value, "The covariance matrices are not equal"))
} else {
  print(c("p-value =", p_value, " The covariance matrices are equal"))
}

```
We will now try to find a way to classify our wines between red and white based on their chemical properties and composition and on their quality score.

To chose between a linear or quadratic discriminant analysis approach (lda or qda), we started by checking if the covariance matrices for each groupe (\textit{red} or \textit{white}) were equal. To do so, we performed the following test: $H_0$: $\sum_{white.wine} = \sum_{red.wine}$ against $H_1$: $\sum_{white.wine} \neq \sum_{red.wine}$.
Define the Box's M test statistic $M =-2ln(\Lambda) $ (where $\Lambda$ is the likelihood ration test statistics) and $u = [ \sum_{l=1}^{g}\frac{1}{n_l - 1} -\frac{1}{\sum_{l=1}^{2} (n_l - 1)}][\frac{2p^2+3p-1}{6(p+1)(g-1)}]$ where p=11 (number of predictors) and g = 2 (number of classes), we know that $(1-u)M\overset{H_0}{\sim}\chi_{p(p+1)(g-1)/2}^2 = \chi_{66}^2 $ is our test statistic.
The resulting p-value is 1, thus, we fail to reject $H_0$: the data does not provide evidence going against the equality of the means (with confidence level higher than 95\%). Therefore, a lda model would be more appropriate in this situation. 
### LDA 
```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}
winedata$wine_type <- as.factor(ifelse(winedata$wine_type == "red", 1, 0))
levels(winedata$wine_type)
t = nrow(winedata)*0.8 #train sample size
test_ind = sample(nrow(winedata), t) #test indices

wine_train <- winedata[test_ind, ]

wine_validation <- winedata[-test_ind, ]

```
wine_type is a binary variable and, for analysis purpose, we code wine_type as follows: red = 1, white = 0. This conversion is made within the dataset before fitting the model.
Moreover, to evaluate the performance of our model, we selected a random sample containing 80\% of the wine data that will be used as a training set (wine_train) and the other 20\% will be used as a validation set (wine_validation). This will allow us to use cross validation. 
Because we only need quantitative data for these analysis, we will not consider \textit{quality} as a predictor.
```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}

lda.wine = lda(wine_type ~.,data=wine_train[,c(-12)], CV = T)
cross_tab = table(wine_train$wine_type,lda.wine$class)
diag(prop.table(cross_tab,1))
sum(diag(prop.table(cross_tab)))
```
After building the model, we want to evaluate its performance in two ways. 
First, we want to check that it performs well in predicting the data that have been used to build it (wine_train). When generating the cross validation matrix, we can see that more than 99\% of white wines and more than 98\% of red wines were correctly classified. On average, more than 99\% of all predictions were correct.
Even though these results are very high, we might fear that such good results could be due to overfitting ; that is the reason why we also need to check its performance on the validation set. 

```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}
lda.wine = lda(wine_type ~.,data=wine_train[,c(-12)]);
lda_pred = predict(lda.wine, wine_validation)
cross_tab = table(wine_validation$wine_type,lda_pred$class)
diag(prop.table(cross_tab,1))
sum(diag(prop.table(cross_tab)))

```
Surprisingly, on new data, the lda model performs even better. More than 99\% of the predictions made on both red and white wines are correct and, thus, on average, almost all predictions are also correct.

### QDA 

```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}

qda.wine = qda(wine_type ~.,data=wine_train[,c(-12)], CV = T);

cross_tab = table(wine_train$wine_type,qda.wine$class)
diag(prop.table(cross_tab,1))
sum(diag(prop.table(cross_tab)))

qda.wine = qda(wine_type ~.,data=wine_train[,c(-12)]);

qda_pred = predict(qda.wine, wine_validation)
cross_tab = table(wine_validation$wine_type,qda_pred$class)
diag(prop.table(cross_tab,1))
sum(diag(prop.table(cross_tab)))
```
Even though the linear separation was very satisfying and seemed to be the most appropriate, we might want to check if a quadratic discriminant analysis would also be able to capture the limit between the two categories. 

The performance are still very good: between 98 and 99\% of the predictions are correct on both white and red wines, when the predictions are made on the training dataset and it goes over 99\% with red wine, when the validation set is used.
On average, the qda model is correct 98\% of the time. This is a highly satisfying result too.

Since we have to make a choice between the two, however, we will select the lda model: not only does it performs better (even though the difference is not substancial), it is also simpler and more appropriate with the data. Indeed, we showed that the red_wine and white_wine covariance matrices, which is also a sign that the lda model is a better choice. 
### Cluster Analysis (Jiali)
#### 1. Kmeans
```{r}
names(wine)

table(wine$quality)
table(wine$wine_type)

kmeans_c4_wine = kmeans(wine[,1:11],2,nstart = 100);

table(wine[,13],kmeans_c4_wine$cluster);
```



## c. Canonical Correlation Analysis(Jiali)

```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(CCA);

wine_scale = scale(wine[,1:11]);

# Acidity Levels
acidity_levels <- wine_scale[, c("fixed.acidity", 
                                "volatile.acidity", 
                                "citric.acid", 
                                "pH")]

# Sugar and Sulfur Dioxide
sugar_sulfur_dioxide <- wine_scale[, c("residual.sugar", 
                                      "free.sulfur.dioxide", 
                                      "total.sulfur.dioxide")]

# Other Chemical Properties
other_chemical_properties <- wine_scale[, c("chlorides", 
                                           "density", 
                                           "sulphates", 
                                           "alcohol")]
```
Canonical Correlation Analysis is a multivariate statistical method used to examine the relationships between two sets of variables. By identifying pairs of canonical variates—one for each set—that are maximally correlated, CCA helps to understand how one set of variables might predict or relate to another set.

**Canonical Variables**:
- Canonical variables (or variates) are synthetic variables created as linear combinations of the original variables within each set. These are formulated so that their correlations across the sets are maximized. 
- The canonical variables for each set are derived as follows:
  \[ u_i = a_{1i}x_1 + a_{2i}x_2 + \ldots + a_{mi}x_m \]
  \[ v_i = b_{1i}y_1 + b_{2i}y_2 + \ldots + b_{ni}y_n \]
  where \(a_{ji}\) and \(b_{ji}\) are coefficients optimizing the correlation between \(u_i\) and \(v_i\).

**Selection Criteria**:
- The number of canonical variates generated is the lesser of the number of variables in the two sets. Each pair of variates is orthogonal (independent) to the others within its own set.
- The correlation coefficients (\(r_i\)) between the pairs of canonical variates provide a measure of their association.

**Variable Sets**:

- **Set 1 (Acidity Levels)**: Includes fixed acidity, volatile acidity, citric acid, and pH. These variables were chosen because they represent fundamental aspects of wine's chemical nature that directly affect its taste, stability, and fermentation process.
- **Set 2 (Sugar and Sulfur Dioxide)**: Includes residual sugar, free sulfur dioxide, and total sulfur dioxide. This group is critical for understanding the wine's sweetness, preservation, and anti-oxidative properties.
- **Set 3 (Other Chemical Properties)**: Includes chlorides, density, sulphates, and alcohol. These properties influence the wine's preservation, body, and overall quality.

#### 1. Acidity Levels & Sugar and Sulfur Dioxide
This analysis aims to explore and quantify the relationships between two critical sets of variables in wine data: **Acidity Levels** (comprising fixed acidity, volatile acidity, citric acid, and pH) and **Sugar and Sulfur Dioxide** (including residual sugar, free sulfur dioxide, and total sulfur dioxide).

```{r echo=FALSE, results='hide'}
cc1 = cc(acidity_levels, sugar_sulfur_dioxide);
cc1$cor;
cc1[3:4]
```

**Canonical Correlations**:
The first three pairs of canonical variates yielded the following correlations:

1. **First pair**: 0.55497110 (moderate correlation)
2. **Second pair**: 0.20137538 (low correlation)
3. **Third pair**: 0.03044709 (negligible correlation)

These results indicate how well combinations of acidity-related variables correlate linearly with combinations of sugar and sulfur dioxide-related variables.

**Canonical Coefficients**:

- **Acidity Levels Coefficients (X Coefficients)**:
  - **First Canonical Variate**:
    - Fixed Acidity: -0.7190255
    - Volatile Acidity: -0.3682300
    - Citric Acid: 0.3076262
    - pH: -0.4254979
  - **Further Variates** show varying influence of these acidity variables, reflecting different aspects of their interaction with sugar and sulfur dioxide levels.

- **Sugar and Sulfur Dioxide Coefficients (Y Coefficients)**:
  - **First Canonical Variate**:
    - Residual Sugar: 0.08087317
    - Free Sulfur Dioxide: 0.13898312
    - Total Sulfur Dioxide: 0.85208294
  - **Further Variates** exhibit distinct patterns, indicating more complex and less pronounced relationships.

The analysis indicates a significant but moderate initial correlation, driven primarily by total sulfur dioxide and fixed acidity. This suggests that as acidity levels in wine change, they might be closely linked with changes in sulfur dioxide levels, which play a crucial role in wine preservation and stability.

The weaker correlations observed in the second and third variates suggest that while there are additional relationships between these sets of variables, they are less straightforward and possibly influenced by other factors not captured solely by these measurements.

Canonical Correlation Analysis between Acidity Levels and Sugar and Sulfur Dioxide in wine reveals that there are meaningful but complex interactions between these variables. Understanding these relationships can help in optimizing wine production to enhance both taste and longevity, particularly by managing acidity and preservative levels effectively.

#### 2. Acidity Levels & Other Chemical Properties
The objective of this analysis is to explore and quantify the relationships between two sets of wine-related variables: Acidity Levels (fixed acidity, volatile acidity, citric acid, and pH) and Other Chemical Properties (chlorides, density, sulphates, and alcohol), utilizing Canonical Correlation Analysis (CCA).

```{r echo=FALSE, results='hide'}
cc2 = cc(acidity_levels, other_chemical_properties);
cc2$cor;
cc2[3:4]
```

**Canonical Correlations**:
The analysis produced the following canonical correlations between the variate pairs:

1. **First pair**: 0.68820274 (strong correlation)
2. **Second pair**: 0.25636970 (moderate correlation)
3. **Third pair**: 0.16665241 (weak correlation)
4. **Fourth pair**: 0.01562419 (negligible correlation)

These correlations reveal the varying degrees of linear relationships between the constructed variates of the two sets.

**Canonical Coefficients**:

- **Acidity Levels Coefficients (X Coefficients)**:
  - **First Canonical Variate**: Predominantly influenced by fixed acidity.
  - **Second Canonical Variate**: Volatile acidity is the major contributor.
  - **Third Canonical Variate**: Led by fixed acidity.
  - **Fourth Canonical Variate**: Citric acid is the dominant variable.

- **Other Chemical Properties Coefficients (Y Coefficients)**:
  - **First Canonical Variate**: Density has the most significant negative influence.
  - **Second Canonical Variate**: Chlorides are the main factor.
  - **Third Canonical Variate**: Density plays a pivotal role.
  - **Fourth Canonical Variate**: Alcohol is the major contributor.

The first canonical variate pair shows a strong correlation, indicating a significant relationship between a combination of acidity level variables (especially fixed acidity) and a combination of other chemical properties (dominantly density). This suggests that aspects such as the body and heaviness of wine, which are influenced by density, are closely linked to its acid content.

Subsequent variate pairs show decreasing correlations, with the fourth pair being almost negligible, indicating limited to no linear relationship for that combination of variables.

This Canonical Correlation Analysis has highlighted significant and varied relationships between acidity levels and other chemical properties of wine. The strongest correlation suggests a pivotal interaction between the wine's acidity and its physical characteristics such as density and alcohol content, which can greatly influence winemaking decisions and wine quality assessments.

#### 3. Sugar and Sulfur Dioxide & Other Chemical Properties

The goal of this analysis is to understand the relationships between two sets of wine-related variables: **Sugar and Sulfur Dioxide** (residual sugar, free sulfur dioxide, total sulfur dioxide) and **Other Chemical Properties** (chlorides, density, sulphates, and alcohol), using Canonical Correlation Analysis.

```{r echo=FALSE, results='hide'}
cc3 = cc(sugar_sulfur_dioxide, other_chemical_properties);
cc3$cor;
cc3[3:4]
```

**Canonical Correlations**:
The canonical correlations for the first three pairs of variates are as follows:

1. **First pair**: 0.735553383 (strong correlation)
2. **Second pair**: 0.470342155 (moderate correlation)
3. **Third pair**: 0.008172205 (negligible correlation)

These values indicate the strength of the linear relationships between the canonical variates from each set.

**Canonical Coefficients**:

- **Sugar and Sulfur Dioxide Coefficients (X Coefficients)**:
  - **First Canonical Variate**: Dominated by residual sugar.
  - **Second Canonical Variate**: Strong influence from total sulfur dioxide.
  - **Third Canonical Variate**: Largely influenced by total sulfur dioxide and negatively by free sulfur dioxide.

- **Other Chemical Properties Coefficients (Y Coefficients)**:
  - **First Canonical Variate**: Primarily driven by density.
  - **Second Canonical Variate**: Notably influenced by alcohol and chlorides.
  - **Third Canonical Variate**: Dominated by chlorides.


The analysis reveals a strong initial correlation suggesting a significant link between the sugar/sulfur dioxide content and the density of the wine. This could suggest that wines with higher residual sugar potentially exhibit higher densities, which might influence sensory qualities like body and sweetness.

Subsequent variates display moderate to negligible correlations, indicating less pronounced relationships for the other combinations of variables. The second variate suggests an interesting interaction between sulfur dioxide content and alcohol levels, possibly indicating how preservation strategies could be related to alcohol content for balance in winemaking.

This Canonical Correlation Analysis highlights important relationships between the sugar and sulfur dioxide properties of wine and its other chemical characteristics, particularly the strong link between wine sweetness and its physical density. These insights can inform more nuanced approaches to balancing these properties in wine production, enhancing both preservation and sensory qualities.

## d. Prediction or classification (Maddie)
## Logistic regression (chapter 11)
```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}

winedata <- rbind(red_wine, white_wine)

winedata$wine_type <- ifelse(winedata$wine_type == "red", 1, 0)
winedata$wine_type <- as.factor(winedata$wine_type)

lg.wine_type = glm(wine_type ~ volatile.acidity + citric.acid 
             + residual.sugar + chlorides + free.sulfur.dioxide + 
               total.sulfur.dioxide + density + sulphates + alcohol 
             + quality, family = binomial(link ="logit"), data = winedata)
lg.wine_type = glm(wine_type ~ ., family = binomial(link ="logit"), data = winedata)
summary(lg.wine_type)


```
As previously stated, wine_type is a categorical variable, which can take the two following values: white or red. 
In order to predict the type of a certain wine based on its chemical composition and properties, we fit a logistic regression model. It is appropriate since wine_type is a binary variable. Because the result of such a model implicitly depends on a probability score, which will be rounded to the closest between 0 and 1, we coded wine_type as follows: red = 1, white = 0. This conversion was made within the dataset before fitting the model.
$$ $$
We have no previously held knowledge or method that would have permitted us to identify which variables were the most influencial in this situation. Therefore, we first fitted a model that contained all possible predictors. The Student tests performed on each of their associated coefficients indicated that the data showed no evidence that the ones associated to \textit{pH} and \textit{fixed.acidity} would not be zero (their p-values were, respectively, 0.1689 and 0.0861, which is higher than 0.05). This implies that these variables do not have a significant impact on the outcome, with 95\% confidence ; thus, they could easily be removed from the model. However, these tests are performed individually for each coefficients, hence, we cannot remove both of them at once, since their impact could be changed by the absence of the other. The p-value for the coefficient associated to \textit{pH} was higher, indicating a lower significance, therefore, it is the one that we chose to remove first.
When the second model was fitted (with all possible predictors but \textit{pH}), fixed.acidity was still not significant (for its coefficient, the Student test's p-value was 0.2985>0.05). We then removed it and fitted a new model without \textit{pH} nor \textit{fixed.acidity}.

Now, note $f(X) = -1645 + 7.10X_2 - 2.83 X_3 - 0.871 X_4 + 24.3 X_5 + 0.0586 X_6 - 0.0522 X_7 + 1636 X_8 + 3.06 X_{10} + 1.56 X{11} + 0.410 X_{12}$. We are trying to predict $Y = X_{13}$, thus, our final model is:
$Y = \mathbb{1}_{f(X)>0.5}$.

### Multivariate Regression
```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}
wine.quant = winedata[,c(-13)]

#BIC backward model selection
step(lm(quality ~ ., data = wine.quant), direction = "backward",
     k=log(nrow(wine.quant)))

#fit the selected model
fit.wine = lm(formula = quality ~ fixed.acidity + volatile.acidity + residual.sugar + 
    free.sulfur.dioxide + total.sulfur.dioxide + density + pH + 
    sulphates + alcohol, data = wine.quant)
summary(fit.wine)
```
We now want to fit a model to predict a wine's quality score $X_{12}$ based on its chemicals properties and composition.
Quality is, in essence a qualitative variable, since it can only take integer values between 0 and 10. However, the range is large enough so we can consider it quantitative. When it comes to prediction, it would even allow for fraction of points, which would be useful in prediction: wine producers would be able to know how far they are from a certain score whereas consumers would know if a wine is in lower or higher range of the said score. 

Considering quality as a qualitative variable is also practical for us to chose a model to fit to our data. Indeed, we will try to predict \textit{quality} with multivariate regression. Since we want the predictors to be only quantitative as well, we removed \textit{wine_type} upfront, since it is binary. Then, we will use the BOC model selection algorithm in order to remove insignificant predictors and select the ones that would maximize the likelihood function associated to $X_{13}$. The BIC penalizes large models, to avoid the risk of overparametrization, which is also useful in our case, since we have 13 parameters.
This time, we want to predict  $Y = X_{12}$
In the end, the chosen model is $Y = 60 + 0.066 X_1 -1.30 X_2 + 0.045 X_4 + 0.0059 X_6 - 0.0025 X_7 - 0.59 X_8+ 0.48 X_9 + 0.74 X_{10} + 0.26 X_{11}$.

After displaying the summary for this model, we confirm that all parameters are indeed significant at level 95\% (from the Student tests) and that, overall, it demonstrate strong performance compared to a constant model (small Fisher test p-value). The adjusted R-squared, however, is low (only 0.2906). We tried improving the model by adding or removing parameters, but we were never able to improve this indicator. 

## e. Hypothesis construction and testing (Jarrad)
### one-way MANOVA
